Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-04-30 overleaf/compiler.tex
Edits across several sections.

\begin{enumerate}
\item Loops are of the form $0 <= i < I$ and bounds are fixed at compile time.
It is a standard restriction in MPC that the bounds must be known at circuit-generation time.
\item Arrays are one-dimensional. $N$-dimensional arrays are linearized and accessed
in row-major order and at this point the programmer is responsible for linearization
and access. (This restriction can be easily lifted.)
\item Array subscrpts are plaintext values as specified by the rule:
\begin{semantics}
    \ntyperule{Array Access}{
%  \Gamma(\y) =  q_\y   \quad\quad \sigma(\f) = q_\f  \quad\quad \Gamma(\x) = q_\x \\
  \Gamma \vdash e : \langle \texttt{plain int} \rangle \quad \Gamma \vdash \texttt{A} : \langle q \; \texttt{list}[\tau] \rangle \quad \tau \in \{ \texttt{int}, \texttt{bool} \}
}{
  \Gamma \vdash \texttt{A[$e$]} : \langle q \; \tau \rangle
}
\end{semantics}
The subscript $e$ is a function of the indices of the enclosing loops.
For read access, the compiler allows an arbitrary such function.
However, it restricts write access to \emph{canonical writes}, i.e., {\sf A[i,j,k] = ...}
where $i$, $j$ and $k$ loop over the three dimensions of {\sf A}.
Write access such as for example {\sf A[i,j+2] = ...} is not allowed.
%\item \ana{Add rules for logical ops}
%\item \ana{Add rules for arithmetic ops}
\item The final restriction involves MUX as expressed by the rule:
\begin{semantics}
    \ntyperule{MUX}{
%  \Gamma(\y) =  q_\y   \quad\quad \sigma(\f) = q_\f  \quad\quad \Gamma(\x) = q_\x \\
  \Gamma \vdash e_1 : \langle q_1 \; \texttt{bool} \rangle \quad \Gamma \vdash e_2 : \langle q_2 \; \tau \rangle \quad \Gamma \vdash e_2 : \langle q_2 \; \tau \rangle \quad \tau \in \{ \texttt{int}, \texttt{bool} \}
}{
  \Gamma \vdash \texttt{MUX($e_1$,$e_2$,$e_3$)} : \langle q_1 \vee q_2 \vee q_3  \; \tau \rangle
}
\end{semantics}
\end{enumerate}
Importantly, the arguments of MUX are restricted to base types, as passing array arguments
is not valid MOTION MPC. \ana{Check this.} This causes a slight inconvenience as we could not write
\begin{pythonn}
if e: A[i] = val
\end{pythonn}
Instead we had to write
\begin{pythonn}
if not(e): val = A[i]
A[i] = val
\end{pythonn}

We note that while the programmer writes
annotations at the level of IMP Source (recall \ana{ref to overview section}), the
annotations propagate through the transformations and are checked at the
level of MPC Source.

>>>>

\begin{enumerate}
\item Loops are of the form $0 <= i < I$ and bounds are fixed at compile time.
It is a standard restriction in MPC that the bounds must be known at circuit-generation time.
\item Arrays are one-dimensional. $N$-dimensional arrays are linearized and accessed
in row-major order and at this point the programmer is responsible for linearization
and access. (This restriction can be easily lifted.)
\item Array subscrpts are plaintext values as specified by the rule:
\begin{semantics}
    \ntyperule{Array Access}{
%  \Gamma(\y) =  q_\y   \quad\quad \sigma(\f) = q_\f  \quad\quad \Gamma(\x) = q_\x \\
  \Gamma \vdash e : \langle \texttt{plain int} \rangle \quad \Gamma \vdash \texttt{A} : \langle q \; \texttt{list}[\tau] \rangle \quad \tau \in \{ \texttt{int}, \texttt{bool} \}
}{
  \Gamma \vdash \texttt{A[$e$]} : \langle q \; \tau \rangle
}
\end{semantics}
The subscript $e$ is a function of the indices of the enclosing loops.
For read access, the compiler allows an arbitrary such function.
However, it restricts write access to \emph{canonical writes}, i.e., {\sf A[$i,j,k$] = ...}
where $i$, $j$ and $k$ loop over the three dimensions of {\sf A}.
Write access such as for example {\sf A[$i,j+2$] = ...} is not allowed.
%\item \ana{Add rules for logical ops}
%\item \ana{Add rules for arithmetic ops}
\item The final restriction involves MUX as expressed by the rule:
\begin{semantics}
    \ntyperule{MUX}{
%  \Gamma(\y) =  q_\y   \quad\quad \sigma(\f) = q_\f  \quad\quad \Gamma(\x) = q_\x \\
  \Gamma \vdash e_1 : \langle q_1 \; \texttt{bool} \rangle \quad \Gamma \vdash e_2 : \langle q_2 \; \tau \rangle \quad \Gamma \vdash e_2 : \langle q_2 \; \tau \rangle \quad \tau \in \{ \texttt{int}, \texttt{bool} \}
}{
  \Gamma \vdash \texttt{MUX($e_1$,$e_2$,$e_3$)} : \langle q_1 \vee q_2 \vee q_3  \; \tau \rangle
}
\end{semantics}
\end{enumerate}
Importantly, the arguments of MUX are restricted to base types, as passing array arguments
is not valid MOTION MPC. \ana{Check this.} This causes a slight inconvenience as we could not write
\begin{pythonn}
if e: A[i] = val
\end{pythonn}
Instead we had to write
\begin{pythonn}
if not(e): val = A[i]
A[i] = val
\end{pythonn}

We note that while the programmer writes
annotations at the level of IMP Source (as in Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source}(a)), the
annotations propagate through the transformations and are checked at the 
level of MPC Source. Recall that the programmer annotates only inputs

### 2022-04-30 overleaf/prob_stmt_np_hardness.tex
Edits across several sections.

As an example, consider the MPC program consisting of the following three sequences, all made up of two distinct $\alpha$-instructions $M_1$ and $M_2$, e.g., a $M_1$ is MUL and $M_2$ is MUX. The right arrow indicates a def-use \emph{dependence}, meaning that the source node must execute before the target node:

@@

The problem of finding a schedule $P$ with a minimal $cost(P)$ is shown to be NP-Hard problem, as it can be reduced to the problem of finding a \emph{shortest common supersequence}, a known NP-Hard problem\cite{Maier1978, Vazirani2010}. The shortest common supersequence problem is as follows: {\it given two or more sequences find the the shortest sequence that contains all of the original sequences.} This can be solved in $O(n^k)$ time, where $n$ is the cardinality of the longest sequence and $k$ is the number of sequences. For our problem $n$ is the maximum length of a sequence and $k$ is the number of instructions of sequences. We can immediately see that the optimal schedule is the shortest schedule, since the shortest schedule minimizes the fixed cost while the variable cost remains the same.

>>>>

As an example, consider the MPC program consisting of the following three sequences, all made up of two distinct $\alpha$-instructions $M_1$ and $M_2$, e.g., $M_1$ is MUL and $M_2$ is MUX. The right arrow indicates a def-use \emph{dependence}, meaning that the source node must execute before the target node:

@@

The problem of finding a schedule $P$ with a minimal $cost(P)$ is shown to be NP-Hard problem, as it can be reduced to the problem of finding a \emph{shortest common supersequence}, a known NP-Hard problem\cite{Maier1978, Vazirani2010}. The shortest common supersequence problem is as follows: {\it given two or more sequences find the the shortest sequence that contains all of the original sequences.} This can be solved in $O(n^k)$ time, where $n$ is the cardinality of the longest sequence and $k$ is the number of sequences. %For our problem $n$ is the maximum length of a sequence and $k$ is the number of instructions of sequences. 
We can immediately see that the optimal schedule is the shortest schedule, since the shortest schedule minimizes the fixed cost while the variable cost remains the same.

### 2022-04-30 overleaf/vectorization.tex
Edits across several sections.

\begin{itemize}
\item same-level edge $X\rightarrow Y$ where $X$ and $Y$ are in the same loop nest, say $i,j,k$. E.g., the def-use edge 9 to 10 in the Biometric MPC Source in Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source} is a same-level edge. A same-level edge can be a back-edge in which case a $\phi$ node is the target of the edge.
E.g., 15 to 4 in Biometric is a same-level back-edge.
\item outer-to-inner $X\rightarrow Y$ where $X$ is in an outer loop nest, say $i$, and $Y$ is in an inner one, say $i,j,k$. E.g., 1 to 4 in Biometric forms is an outer-to-inner edge.
%A $\phi$-node can be a source or a target of an outer-to-inner forward edge.
\item inner-to-outer $X\rightarrow Y$ where $X$ is a $phi$-node in an inner loop nest, $i,j,k$, and $Y$ is in the enclosing loop nest $i,j$. E.g., the def-use from 8 to 12 gives rise to an inner-to-outer edge.
%E.g. $\texttt{sum}_0 = \phi(\texttt{sum}_1, 0)$ to $\texttt{c} = \mathit{CMP}(\texttt{sum}_0,\texttt{min}_0)$ is an inner-to-outer edge.
An inner-to-outer edge can be a back-edge as well, in which case both $X$ and $Y$ are phi-nodes with the source $X$ in a loop nested into $Y$'s loop (not necessarily immediately).

@@

\subsubsection{Overview}

An important component of our algorithm is the scalar expansion to the corresponding loop dimensionality.
For example, {\sf d = S[i*D+j] - C[j]} equiv. to {\sf d = S[i,j] - C[j]}, which gave rise to {\sf N*D} subtraction operations in the sequential schedule,
is lifted. The argument arrays {\sf S} and {\sf C} are lifted and the scaler {\sf d} is lifted: {\sf d[i,j] = S[i,j] - C[i,j]}.
The algorithm then detects that the statement can be vectorized.

There are three kinds of arrays.
\begin{itemize}

@@

\subsubsection{Raise Dimension and Drop Dimension}

There are two conceptual versions of $\mathit{raise\_dim}$. One reshapes an arbitrary access into a canonical read access in the corresponding loop. 
%The signature of $\mathit{raise\_dim}$ is as follows. 
It takes the original array, the access pattern function $f(i,j,k)$ in loop nest $i,j,k$ and the loop bounds
$((i:I),(j:J),(k:K))$:
\[ \mathit{raise\_dim}({\sf A, f(i,j,k), ((i:I),(j:J),(k:K))}) \]
It produces a new 3-dimensional array {\sf A'} by iterating over $i,j,k$ and setting each element of  {\sf A'} as follows:
\[ {\sf A'[i,j,k] =  A[f(i,j,k)]} \]
The end result is that uses of {\sf [A[f(i,j,k)]} in loop nest $i,j,k$ are replaced with canonical read-accesses to {\sf A'[i,j,k]}
that can be vectorized. In the running Biometric example, {\sf C'} = $\mathit{raise\_dim}${\sf (C, j, (i:N,j:D))} lifts the
1-dimensional array {\sf C} into a 2-dimensional array. The $i,j$ loop now accesses {\sf C'} in the canonical way, {\sf C'[i,j]}.
Similarly, {\sf S'} = $\mathit{raise\_dim}${\sf (S, i*D+j, (i:N,j:D))} tries to lift {\sf S}, but the operation turns into a no-op
because {\sf S} is already a 2-dimensional array and the read access is canonical.

The other version of $\mathit{raise\_dim}$ lifts a lower-dimension array into a
higher-dimension for access in a nested loop. It is necessary when processing outer-to-inner dependences. 
Here {\sf A} is an $i$-array and raise dimension adds two additional dimensions:
\[ \mathit{raise\_dim}({\sf A}, {\sf (j:J,k:K)}) \]
This version is reduced to the above version by adding the access pattern function, which is just $i$:
\[ \mathit{raise\_dim}({\sf A, i, (j:J,k:K)}) \]

>>>>

\begin{itemize}
\item same-level edge $X\rightarrow Y$ where $X$ and $Y$ are in the same loop nest, say $i,j,k$. E.g., the def-use edge 9 to 10 in the Biometric MPC Source in Listing~\ref{tab:source_and_MPC_source_and_optimized_MPC_source} is a same-level edge. A same-level edge can be a back-edge in which case a $\phi$ node is the target of the edge.
E.g., 15 to 4 in Biometric is a same-level back-edge.
\item outer-to-inner $X\rightarrow Y$ where $X$ is in an outer loop nest, say $i$, and $Y$ is in an inner one, say $i,j,k$. E.g., 1 to 4 in Biometric forms is an outer-to-inner edge.
%A $\phi$-node can be a source or a target of an outer-to-inner forward edge.
\item inner-to-outer $X\rightarrow Y$ where $X$ is a $phi$-node in an inner loop nest, $i,j,k$, and $Y$ is in the enclosing loop nest $i,j$. E.g., the def-use from 8 to 12 gives rise to an inner-to-outer edge.
%E.g. $\texttt{sum}_0 = \phi(\texttt{sum}_1, 0)$ to $\texttt{c} = \mathit{CMP}(\texttt{sum}_0,\texttt{min}_0)$ is an inner-to-outer edge.
An inner-to-outer edge can be a back-edge as well, in which case both $X$ and $Y$ are $\phi$-nodes with the source $X$ in a loop nested into $Y$'s loop (not necessarily immediately).

@@

%\subsubsection{Overview}

An important component of our algorithm is the scalar expansion to the corresponding loop dimensionality, 
which is necessary to expose opportunities for vectorization.
For example, {\sf d = S[i*D+j] - C[j]} equiv. to {\sf d = S[i,j] - C[j]}, which gave rise to {\sf N*D} subtraction operations in the sequential schedule,
is lifted. The argument arrays {\sf S} and {\sf C} are lifted and the scaler {\sf d} is lifted: {\sf d[i,j] = S[i,j] - C[i,j]}.
The algorithm then detects that the statement can be vectorized.

\paragraph{Arrays} There are three kinds of arrays.
\begin{itemize}

@@

%\subsubsection{Raise Dimension and Drop Dimension}

\paragraph{Raise dimension} 
There are two conceptual versions of $\mathit{raise\_dim}$. One reshapes an arbitrary access into a canonical read access in the corresponding loop. 
%The signature of $\mathit{raise\_dim}$ is as follows. 
It takes the original array, the access pattern function $f(i,j,k)$ in loop nest $i,j,k$ and the loop bounds
$((i:I),(j:J),(k:K))$:
\[ \mathit{raise\_dim}({\sf A, f(i,j,k), ((i:I),(j:J),(k:K))}) \]
It produces a new 3-dimensional array {\sf A'} by iterating over $i,j,k$ and setting each element of  {\sf A'} as follows:
\[ {\sf A'[i,j,k] =  A[f(i,j,k)]} \]
The end result is that uses of {\sf [A[f(i,j,k)]} in loop nest $i,j,k$ are replaced with canonical read-accesses to {\sf A'[i,j,k]}
that can be vectorized. In the running Biometric example, {\sf C'} = $\mathit{raise\_dim}${\sf (C, j, (i:N,j:D))} lifts the
1-dimensional array {\sf C} into a 2-dimensional array. The $i,j$ loop now accesses {\sf C'} in the canonical way, {\sf C'[i,j]}.
Similarly, {\sf S'} = $\mathit{raise\_dim}${\sf (S, i*D+j, (i:N,j:D))} tries to lift {\sf S}, but the operation turns into a no-op
because {\sf S} is already a 2-dimensional array and the read access is canonical.

\paragraph{Drop dimension} 
The other version of $\mathit{raise\_dim}$ lifts a lower-dimension array into a
higher-dimension for access in a nested loop. It is necessary when processing outer-to-inner dependences. 
Here {\sf A} is an $i$-array and raise dimension adds two additional dimensions:
\[ \mathit{raise\_dim}({\sf A}, {\sf (j:J,k:K)}) \]
This version is reduced to the above version by adding the access pattern function, which is just $i$:
\[ \mathit{raise\_dim}({\sf A, i, (j:J,k:K)}) \]
